\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-26 paper.qmd
Added total number of bins section

The number of hexagonal bins in the hexagonal grid has a considerable impact on the construction of the 2D model. This is because it is the initial step in building the 2D model. The hexagonal grid with the chosen total number of bins must be able to capture the structure of the NLDR data. If the number of bins is too low, the model may not be able to capture the structure of the NLDR data effectively (see @fig-binsize (a)), while if there are too many bins, it may result in over-fitting the individual points of the NLDR data (see @fig-binsize (c)). Therefore, it is important to determine an appropriate number of bins to build an effective model.

@@

Furthermore, the total number of bins is determined by the number of bins along the x-axis and y-axis (see @eq-equation8). To calculate the effective total number of bins, candidate values are selected based on the range between the minimum and approximate maximum number of bins along the x and y axes. The minimum number of bins along each axis is set to $1$, while the maximum number is estimated by taking the square root of the NLDR data point count. The analysis evaluates the MSE across varying total bin counts within this range, covering the minimum to maximum values along both axes.

The analysis focuses on exploring the effect of total number of bins on MSE. It considers MSE across different total number of bins, covering the range calculated from the minimum to maximum values along both x and y axes.

Typically, the MSE plot tends to have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or levesl off, indicating that a higher number of bins generates a smaller error. This MSE plot is useful in determining the effective number of bins required to construct the 2D model. @fig-diagnosticpltScurve has created a graph that shows how the Mean Squared Error (MSE) changes depending on the number of bins assigned in the hexagonal grid for UMAP applied to the S-curve dataset. As per the rule of thumb mentioned earlier, the effective number of bins is determined to be $84$, with $6$ bins along the x-axis and $14$ bins along the y-axis. Furthermore, @fig-modelScurve shows the 2D model constructed using UMAP with the selected total number of bins and how the model overlay on high-D data.

@@

## To draw with AIC
# aic_plot <- ggplot(mse_df, aes(x = num_bins, y = aic
# )) +
#   geom_point() +
#   geom_line() +
#   geom_vline(xintercept = 84, linetype="solid",
#                 color = "red", size=0.8, alpha = 0.5) +
#   theme_light() +
#   theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
#         axis.title = element_text(size = 7),
#         axis.text = element_text(size = 7)) +
#   #scale_colour_manual(values = c("#377eb8")) +
#   ylab("AIC") +
#   xlab("total number of bins")f
## Effective number of bins along x-axis

@@

After establishing the hexagonal grid with an appropriate number of bins, some hexagonal bins may have few or no data points within them. To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagonal bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid (see @eq-equationp2). Next, bins with a standard count less than a certain benchmark value are removed. The following steps will help in determining a suitable value for removing low-density hexagons:

@@

Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven 2D model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density ones. If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the 2D model.

>>>>

The number of hexagonal bins in the hexagonal grid has a considerable impact on the construction of the 2D model, serving as the initial step in building the 2D model. The chosen total number of bins must effectively capture the structure of the NLDR data. If the number of bins is too low, the model may not be able to capture the structure of the NLDR data effectively (see @fig-binsize (a)), while if there are too many bins, it may result in over-fitting the individual points of the NLDR data (see @fig-binsize (c)). Therefore, it is important to determine an appropriate number of bins to construct an effective model.

@@

Furthermore, the total number of bins is calculated based on the number of bins along the x-axis and y-axis, as shown in @eq-equation8. To determine the effective total number of bins, candidate values are selected based on the range between the minimum and approximate maximum number of bins along the x and y axes. The minimum number of bins along each axis is set to $1$, while the maximum number is estimated by taking the square root of the NLDR data points. The analysis evaluates the MSE across varying total number of bins within this range, covering the minimum to maximum values along both axes (see @fig-diagnosticpltScurve).

@@

@@

After establishing the hexagonal grid with an appropriate number of bins, some hexagonal bins may have few or no data points within them (see @fig-rmlowdenshex (a)). To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagonal bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid (see @eq-equationp2). Next, bins with a standard count less than a certain benchmark value are removed (see @fig-rmlowdenshex (b)). The following steps will help in determining a suitable value for removing low-density hexagons:

@@

Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven 2D model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density ones (see @fig-rmlowdenshex (c)). If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the 2D model.

### 2024-03-26 paper.tex
Added total number of bins section

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, frame hidden, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the 2D model. This is because it is the
initial step in building the 2D model. The hexagonal grid with the
chosen total number of bins must be able to capture the structure of the
NLDR data. If the number of bins is too low, the model may not be able
to capture the structure of the NLDR data effectively (see
Figure~\ref{fig-binsize} (a)), while if there are too many bins, it may
result in over-fitting the individual points of the NLDR data (see
Figure~\ref{fig-binsize} (c)). Therefore, it is important to determine
an appropriate number of bins to build an effective model.

@@

Furthermore, the total number of bins is determined by the number of
bins along the x-axis and y-axis (see Equation~\ref{eq-equation8}). To
calculate the effective total number of bins, candidate values are
selected based on the range between the minimum and approximate maximum
number of bins along the x and y axes. The minimum number of bins along
each axis is set to \(1\), while the maximum number is estimated by
taking the square root of the NLDR data point count. The analysis
evaluates the MSE across varying total bin counts within this range,
covering the minimum to maximum values along both axes.

The analysis focuses on exploring the effect of total number of bins on
MSE. It considers MSE across different total number of bins, covering
the range calculated from the minimum to maximum values along both x and
y axes.

Typically, the MSE plot tends to have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope gradually declines or levesl off, indicating that
a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} has created a graph
that shows how the Mean Squared Error (MSE) changes depending on the
number of bins assigned in the hexagonal grid for UMAP applied to the
S-curve dataset. As per the rule of thumb mentioned earlier, the
effective number of bins is determined to be \(84\), with \(6\) bins
along the x-axis and \(14\) bins along the y-axis. Furthermore,
Figure~\ref{fig-modelScurve} shows the 2D model constructed using UMAP
with the selected total number of bins and how the model overlay on
high-D data.

@@

After establishing the hexagonal grid with an appropriate number of
bins, some hexagonal bins may have few or no data points within them. To
ensure comprehensive coverage of the NLDR data, it is necessary to
select hexagonal bins with a considerable number of data points. This
involves calculating the number of points within each hexagon. Then, the
standard count is computed by dividing the number of points within each
hexagon by the maximum number of points in the grid (see
Equation~\ref{eq-equationp2}). Next, bins with a standard count less
than a certain benchmark value are removed. The following steps will
help in determining a suitable value for removing low-density hexagons:

@@

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven 2D model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density ones. If neighboring
bins also show low counts, only those bins will be removed. The
remaining bins are used to construct the 2D model.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the 2D model, serving as the initial step
in building the 2D model. The chosen total number of bins must
effectively capture the structure of the NLDR data. If the number of
bins is too low, the model may not be able to capture the structure of
the NLDR data effectively (see Figure~\ref{fig-binsize} (a)), while if
there are too many bins, it may result in over-fitting the individual
points of the NLDR data (see Figure~\ref{fig-binsize} (c)). Therefore,
it is important to determine an appropriate number of bins to construct
an effective model.

@@

Furthermore, the total number of bins is calculated based on the number
of bins along the x-axis and y-axis, as shown in
Equation~\ref{eq-equation8}. To determine the effective total number of
bins, candidate values are selected based on the range between the
minimum and approximate maximum number of bins along the x and y axes.
The minimum number of bins along each axis is set to \(1\), while the
maximum number is estimated by taking the square root of the NLDR data
points. The analysis evaluates the MSE across varying total number of
bins within this range, covering the minimum to maximum values along
both axes (see Figure~\ref{fig-diagnosticpltScurve}).

@@

After establishing the hexagonal grid with an appropriate number of
bins, some hexagonal bins may have few or no data points within them
(see Figure~\ref{fig-rmlowdenshex} (a)). To ensure comprehensive
coverage of the NLDR data, it is necessary to select hexagonal bins with
a considerable number of data points. This involves calculating the
number of points within each hexagon. Then, the standard count is
computed by dividing the number of points within each hexagon by the
maximum number of points in the grid (see Equation~\ref{eq-equationp2}).
Next, bins with a standard count less than a certain benchmark value are
removed (see Figure~\ref{fig-rmlowdenshex} (b)). The following steps
will help in determining a suitable value for removing low-density
hexagons:

@@

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven 2D model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density ones (see
Figure~\ref{fig-rmlowdenshex} (c)). If neighboring bins also show low
counts, only those bins will be removed. The remaining bins are used to
construct the 2D model.
